---
title: "Using RFAssetPricing: Mixed-Effects Random Forests for Asset Pricing"
author: "Ho Yin (Vic) Chien"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using RFAssetPricing: Mixed-Effects Random Forests for Asset Pricing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

```
## Introduction

The RFAssetPricing package implements various random forest-based methods for longitudinal data analysis in asset pricing. It provides four main model types:

1. MERF (Mixed Effects Random Forest)
2. SMERF (Stochastic Mixed Effects Random Forest)
3. MERT (Mixed Effects Random Tree)
4. SMERT (Stochastic Mixed Effects Random Tree)

These models combine the flexibility of random forests with the ability to handle longitudinal data structure through mixed effects and stochastic processes.

## Model Structure

The general model structure is:


$Y_{i,t} = f(X_{i,t}) + Z_i + \omega(\tau_{i,t}) + \epsilon_{i,t}$
 
where:

 - $Y_{i,t}$ is the output at time $t$ for firm $i$

 - $X_{i,t}$ is a vector of of predictors

 - $Z_i$ represents scalar random effect for firm $i$
 
 - $\tau_{i,t}$ is the measurement time for observation $i,t$

 - $\omega(\tau_{i,t})$ is a stochastic process (for S-variants)

 - $\epsilon_{i,t}$ is the residual error


We are going to learn $f$ using CART/Random forests, and form predictions of $\omega(\tau_{i,t})$ and $Z_i$.


## Basic Usage
First, let's load the package:
```{r}
library(RFAssetPricing)
```


## Simulated Example
We'll start with a simulated dataset to demonstrate the package functionality:

```{r,echo = TRUE}
library(LongituRF)

set.seed(123)
# Generate sample data
sim_data <- DataLongGenerator(n = 50, p = 6, G = 6)

# Examine data structure
str(sim_data)
```

The generated data includes:

 - X: Fixed effects predictors
 
 - Y: Response variable
 
 - Z: Random effects design matrix
 
 - id: Firm identifiers
 
 - time: Time points

## Fitting Models

Let's fit both standard and stochastic variants:
```{r,echo = TRUE}
# Fit MERF model
merf_model <- train_APRF(
    model = "merf",
    train_X = sim_data$X,
    train_Y = sim_data$Y,
    train_time = sim_data$time,
    train_id = sim_data$id,
    n_iter = 50  # Reduced for vignette compilation
)

# Fit SMERF model with Brownian Motion
smerf_model <- train_APRF(
    model = "smerf",
    train_X = sim_data$X,
    train_Y = sim_data$Y,
    train_time = sim_data$time,
    train_id = sim_data$id,
    sto = "BM",
    n_iter = 50
)
```
## Model Comparison

Let's compare the predictions from both models:

```{r,echo = TRUE}
# Get predictions
merf_pred <- predict(merf_model, 
                    X = sim_data$X,
                    Z = sim_data$Z,
                    id = sim_data$id,
                    time = sim_data$time)

smerf_pred <- predict(smerf_model, 
                     X = sim_data$X,
                     Z = sim_data$Z,
                     id = sim_data$id,
                     time = sim_data$time)

# Calculate RMSE
merf_rmse <- sqrt(mean((merf_pred - sim_data$Y)^2))
smerf_rmse <- sqrt(mean((smerf_pred - sim_data$Y)^2))

# Print results
cat("RMSE Comparison:\n")
cat("MERF:", merf_rmse, "\n")
cat("SMERF:", smerf_rmse, "\n")

```

## Visualization

Let's plot the SMERF model predictions for the 6 individuals versus the ground truth

```{r,echo = TRUE}
# Visualize predictions for a few firms
library(ggplot2)

# Create plotting data
plot_data <- data.frame(
    Time = rep(sim_data$time, 3),
    Value = c(sim_data$Y, merf_pred, smerf_pred),
    Type = factor(rep(c("Actual", "MERF", "SMERF"), 
                     each = length(sim_data$Y))),
    Firm = factor(rep(sim_data$id, 3))
)

# Plot first 6 firms
ggplot(subset(plot_data, Firm %in% levels(Firm)[1:6]), 
       aes(x = Time, y = Value, color = Type)) +
    geom_line() +
    facet_wrap(~Firm) +
    theme_minimal() +
    labs(title = "Model Comparisons",
         y = "Response",
         color = "Type")
```

## Convergence Diagnostics
We can examine the convergence of the estimation procedure:

```{r,  fig.show='hold'}
# Plot log-likelihood evolution
plot(smerf_model$Vraisemblance, type = "l",
     xlab = "Iteration", ylab = "Log-likelihood",
     main = "Convergence Plot")
```



## References
1. Breiman, L. (2001). Random forests. Machine learning, 45:5–32.

2. Capitaine, L., Genuer, R., and Thi´ebaut, R. (2021). Random forests for high-dimensional longitudinal data. Statistical methods in medical research, 30(1):166–184.

3. Hajjem, A., Bellavance, F., and Larocque, D. (2011). Mixed effects regression trees for clustered data. Statistics & probability letters, 81(4):451–459.

4. Hajjem, A., Bellavance, F., and Larocque, D. (2014). Mixed-effects random forest for clustered data. Journal of Statistical Computation and Simulation, 84(6):1313–1328.

5. Sela, R. J. and Simonoff, J. S. (2012). Re-em trees: a data mining approach for longitudinal and clustered data. Machine learning, 86:169–207.

